Power and Pitfalls of Generic Smart Contracts

### 2022-03-02 cluster_computing/001_abstract.tex
title, abstract and introduction rewriting + biblio fix

Generics are a powerful feature of programming languages that allows one
to write highly reusable code.
%
More specifically, they are based on the use of type placeholders in order
to produce parametrized code, that can be instantiated for each
concrete type provided for them. 
%
In many programming languages, such as Java, they are implemented by
\emph{erasure}\ie replaced by their upper bound type during compilation into bytecode.
This is safe at source level, since the compiler takes care of checking that
types are correct, before erasure. However, the types of the generated bytecode
are erased and consequently weaker. In a permissionless blockchain, where
every user can call the bytecode of smart contracts installed by other users,
these weaker types pose a risk of attack.
%
Namely, this paper originated from a real security issue that we found
while using generics for writing
smart contracts that implement
\emph{shared entities} (such as a company shared by its shareholders),
for the Hotmoka blockchain, in a subset of Java called Takamaka.
%
The considered case study is particularly important since
the validators' set of the blockchain itself is
a special case of shared entities.
The analysis shows that the power of generics comes at the risk of
allowing a validator of an illegal type, that exploits the work of another
validator in order to earn validation rewards, an actual attack
to the validators' set. This paper proposes a solution
that forces the compiler to generate more precise type information than
that generated by erasure.

>>>>

Smart contracts are pieces of code that are deployed and executed in the context of a blockchain infrastructure in order automatically enforce some effects when particularly events occur. The writing of smart contracts is a complex and critical activity which can benefit from the use of high-level features of programming languages, like generics. In many programming languages, such as Java, generics are implemented by \emph{erasure}\ie replaced by their upper bound type during compilation into bytecode. This is safe at source level, since the compiler takes care of checking that types are correct, before erasure. However, the types of the generated bytecode are erased and consequently weaker. In a permissionless blockchain, where every user can call the bytecode of smart contracts installed by other users, these weaker types pose a risk of attack. In this paper we start from a real security issue that we found while using generics for writing smart contracts that implement \emph{shared entities} for the Hotmoka blockchain, and we propose a possible solution based on an appropriate code rewriting.

### 2022-03-02 cluster_computing/010_introduction.tex
title, abstract and introduction rewriting + biblio fix

Blockchains exploit the redundant, concurrent execution of the same
transactions on a decentralized network of many machines,
in order to enforce their execution in accordance with
a set of predefined rules. Namely, blockchains make it hard, for a single machine,
to disrupt the semantics of the transactions or their ordering: a misbehaving single machine
gets immediately put out of consensus and isolated. Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}
has been the first blockchain's success story. Here
transactions are programmed in a non-Turing complete bytecode language,
almost exclusively used to implement transfers of units of coins between \emph{accounts}.

A few years after Bitcoin, another blockchain, called
Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming
transactions in an actual, imperative and Turing-complete programming language, called Solidity.
Solidity's code is organized in \emph{smart contracts}, that can be seen as
objects that control money. A smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}.
Ethereum's transactions can hence execute much more than coin transfers. Namely,
they run object constructors and methods, which results in a sort
of \emph{world computer} that persists the same objects inside the memory of all the
computers composing the blockchain's network.

In Solidity's bytecode,
non-primitive values are referenced through a very general
\<address> type. For instance, a Solidity method
\<child(Person p, uint256 n) returns Person> actually compiles
into \<child(address p, uint256 n) returns address>, losing most
type information~\cite{CrafaPZ19}.
Since, at run time, it is the bytecode that gets executed,
everything can be passed for \<p>, not just a \<Person> instance, as illustrated in Fig.\,\ref{figure.solidity_problem}.
The compiler cannot even enforce strong typing
by generating defensive type instance checks and casts, because
values are unboxed in Ethereum: they have no attached
type information at run time,
they are just numerical \emph{addresses}.
It follows that inside the \<child> method, an eventual call to a \<Person>'s method
on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>.
In other words, Solidity is not strongly typed.
Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed
to another method, such as on \<p> passed to \<child>, since an attacker can pass crafted
objects for \<p>, with arbitrary implementations for their methods,
which can result in the unexpected execution of
dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that
costed millions of dollars.

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/solidity_problem}
\caption{Example of possible problem that can occur in Solidity due to the absence of a strong typing mechanism.}
\label{figure.solidity_problem}
\end{figure}

Strong typing is one of the reasons that push towards the adoption
of \emph{traditional} programming languages for smart contracts. For instance,
the Cosmos blockchain~\cite{cosmos} uses Go. The
Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
Hyperledger~\cite{hyperldeger} allows Go and Java.
Another reason is the availability of modern
language features, that are missing in Solidity,
such as \emph{generics}\ie the possibility of using
type variables. Generics are a powerful and very useful facility for programming
smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. In Java source code, generics are strongly typed, if no \emph{unchecked operations}
are used~\cite{NaftalinW06}, as it will always be the case in this paper.
However, generics might have security issues
at the level of compiled Java code and this paper originated from a
real issue that we found in our code.

@@

This paper is a revised and extended version of~\cite{BeniniGMS21}.
In comparison to that paper, Sec.~\ref{sec:takamaka} and
Sec.~\ref{sec:validators} are new; eight figures have been added;
all other sections have been expanded.

>>>>

In recent years the Blockchain technology has rapidily emerged as a powerfull tool for supporting the development of many and innovative services and infrastructures. Blockchain-enabled applications are spreading across diverse sectors such as supply chain, business, healthcare, IoT, privacy, and data management \cite{blockchain_applications}. A blockchain is essentially a \textit{distributed ledger}, namely a database replicated across different locations and synchronized by multiple independent participants. Blockchains exploit the redundant, concurrent execution of the same transactions on a decentralized network of many machines, in order to enforce their execution in accordance with a set of predefined rules. Namely, blockchains make it hard, for a single machine, to disrupt the semantics of transactions or their ordering: a misbehaving single machine gets immediately put out of consensus and isolated. 
%
\begin{figure*}[th]
\centering
\includegraphics[width=0.8\linewidth]{figures/solidity_problem}
\caption{Example of possible problem that can occur in Solidity due to the absence of a strong typing mechanism.}
\label{figure.solidity_problem}
\end{figure*}
%
The key innovation introduced by this technology is a mechanism  able to reach an emergent agreement about a global state without the need for a central authority. Morever, another peculiarity is that the consensus is not explicit, because there is not a fixed moment when it occurs.

Bitcoin~\cite{Nakamoto08,book-mastering-bitcoin}  has been the first blockchain's success story. Here transactions are programmed in a non-Turing complete bytecode language, almost exclusively used to implement transfers of units of coins between \emph{accounts}, providing a totally decentralized P2P digital cash system based on a distributed public ledger. 
%
A few years after Bitcoin, another blockchain, called Ethereum~\cite{Buterin13,AntonopoulosW18}, introduced the possibility of programming transactions in an actual, imperative and Turing-complete programming language, called Solidity. The major innovation of Ethereum is the contrustruction through its nodes of a distributed \emph{world computer} that can run general purpose code. Indee, if the term ``distributed ledger'' is usually used to describe blockchains like Bitcoin, Ethereum is often defined as a ``distributed state machine''. Solidity's code is organized in \emph{smart contracts}, namely pieces of code which are stored in the blockchain and are executed when a particular event occurs, e.g. when a transaction is scheduled. From a theoretical point of view, a smart contract is essentially an agreement between two or more parties that can be automatically enforced without the need for a trustworthy intermediary~\cite{ebp}. Through smart contracts, Ethereum's transactions can hence execute much more than coin transfers. In this case the global shared state is given by a set of objects which are persisted and manipulated in the same way by all nodes in the blockchain through the execution of the same object constructors and methods.

The world computer built by Ethereum is known as Ethereum Virtual Machine (EVM) which is the platform where accounts and smart contracts live and are executed.
Solidity is high-level programming language and smart contracts need to be compiled into bytecode to be executed inside the EVM. At this regards, we can observe that in Solidity's bytecode, non-primitive values are referenced through a very general \<address> type. For instance, a Solidity method \<child(Person p, uint256 n) returns Person> actually compiles into \<child(address p, uint256 n) returns address>, losing most type information~\cite{CrafaPZ19}. Since, at run time, it is the bytecode that gets executed, everything can be passed for \<p>, not just a \<Person> instance, as illustrated in Fig.\,\ref{figure.solidity_problem}.
The compiler cannot even enforce strong typing by generating defensive type instance checks and casts, because values are unboxed in Ethereum: they have no attached type information at run time, they are just numerical \emph{addresses}. It follows that inside the \<child> method, an eventual call to a \<Person>'s method on \<p> might actually execute any arbitrary code, if \<p> is not a \<Person>. In other words, Solidity is not strongly typed. Consequently, it is highly discouraged, in Solidity, to call methods on parameters passed to another method, such as on \<p> passed to \<child>, since an attacker can pass crafted objects for \<p>, with arbitrary implementations for their methods, which can result in the unexpected execution of dangerous code. This actually happened in the case of the infamous DAO hack~\cite{dao16}, that costed millions of dollars.

Strong typing is one of the reasons that push towards the adoption
of \emph{traditional} programming languages for smart contracts. For instance,
the Cosmos blockchain~\cite{cosmos} uses Go. The
Hotmoka blockchain~\cite{hotmoka} uses a subset of Java
for smart contracts, called Takamaka~\cite{Spoto19,Spoto20}.
Hyperledger~\cite{hyperldeger} allows Go and Java.
Another reason is the availability of modern
language features, that are missing in Solidity,
such as \emph{generics}\ie the possibility of using
type variables. Generics are a powerful and very useful facility for programming
smart contracts, since they allow one to personalize the behaviour of such contracts and partially overcome their inherent incompleteness~\cite{ebp}. 
%
Through the use of generics, it is possible to provide to users a set of predefined contract templates which can be extended and specialized by users with lower programming skills, but higher knowledge about the specific application domain. They are based on the use of type placeholders in order
to produce parametrized code, that can be instantiated for each
concrete type provided for the placeholders.
%
However, strong typing and generics are two intertwined language features that have to be carefully considered when smart contracts are implemented and their bytecode is subsequently deployed on a blockchain. For instance, in Java source code, generics are strongly typed, if no \emph{unchecked operations} are used~\cite{NaftalinW06}, as it will always be the case in this paper.
However, generics might have security issues at the level of compiled Java code and this paper originated from a real issue that we found in our code.

@@

This paper is a revised and extended version of~\cite{BeniniGMS21}.
In comparison to that paper, Sec.~\ref{sec:takamaka} and
Sec.~\ref{sec:validators} are new; while all other sections have been expanded and the enriched with several explanatory figures.

### 2022-03-02 cluster_computing/017_takamaka.tex
latex style
title, abstract and introduction rewriting + biblio fix

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=12.2cm]{entities-hierarchy.png}
  \end{center}
  \caption{The hierarchy of Takamaka classes that implement accounts, shared entities and validators. Their source code can be found inside the Java project \textsf{\url{https://github.com/Hotmoka/hotmoka/tree/master/io-takamaka-code}}.}\label{fig:hierarchy-entities}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/takamaka_payable_contract}
\caption{Exemplification of the Takamaka persistent objects stored in the blockchain.}
\label{figure.takamaka_payable_contract}
\end{figure}

@@

  public TendermintED25519Validator(String publicKey) {
    super(publicKey);

    MessageDigest sha256 = MessageDigest.getInstance("SHA-256");
    sha256.update(Base64.getDecoder().decode(publicKey()));
    this.id = bytesToHex(sha256.digest()).substring(0, 40);
  }

@@

Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
Both proof of work and proof of stake can be used, for instance. In particular, if proof of stake
is used, then each validator node of the blockchain must specify a
\<io.takamaka.code.governance.Validator> object, that plays the role of the
banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
It is a special externally owned account, with an extra \<id> method that provides the
identifier of the validator node inside the blockchain network. This identifier depends from the
specific network.
For instance, the subclass
\<io.takamaka.code.governance.tendermint.TendermintED25519Validator>
implements \<id> as for the Tendermint blockchain engine, that is,
as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key
(see its code in Fig.~\ref{fig:validator}).

>>>>

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=12.2cm]{entities-hierarchy.png}
  \end{center}
  \caption{The hierarchy of Takamaka classes that implement accounts, shared entities and validators. Their source code can be found inside the Java project \textsf{\url{https://github.com/Hotmoka/hotmoka/tree/master/io-takamaka-code}}.}\label{fig:hierarchy-entities}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figures/takamaka_payable_contract}
\caption{Exemplification of the Takamaka persistent objects stored in the blockchain.}
\label{figure.takamaka_payable_contract}
\end{figure}

@@

  public TendermintED25519Validator
    (String publicKey) {
    super(publicKey);

    MessageDigest sha256 = 
      MessageDigest.getInstance("SHA-256");
    sha256.update(Base64.getDecoder().
      decode(publicKey()));
    this.id = bytesToHex(sha256.digest()).
      substring(0, 40);
  }

@@

Neither the Takamaka language nor the Hotmoka blockchain dictate a specific consensus mechanism.
Both proof of work and proof of stake can be used, for instance. In particular, if proof of stake
is used, then each validator node of the blockchain must specify a
\<io.takamaka.code.governance.Validator> object, that plays the role of the
banking account where the validation rewards of the node get accumulated (see Fig.~\ref{fig:validator}).
It is a special externally owned account, with an extra \<id> method that provides the
identifier of the validator node inside the blockchain network. This identifier depends from the
specific network.
For instance, the subclass
\<io.takamaka.code.governance.tendermint.Tender\-mintED25519Validator>
implements \<id> as for the Tendermint blockchain engine, that is,
as the first $40$ hexadecimal digits of the sha256 digest of the Base64-encoded public key
(see its code in Fig.~\ref{fig:validator}).

### 2022-03-02 cluster_computing/020_shared_entities_with_generics.tex
title, abstract and introduction rewriting + biblio fix

\section{Shared Entities using Generics}\label{sec:shared_entities}

@@

\begin{figure}[ht]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public interface SharedEntity
      <S extends PayableContract, O extends Offer<S>>
      extends SharedEntityView<S> {

@@

    @View public boolean isOngoing() {
      return now() <= expiration;
    }
  }
}
    \end{lstlisting}
  \end{center}
  \caption{A simplified part of our shared entity interface.}\label{fig:shared_entity}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.9\linewidth]{figures/shared_entity}
\caption{Exemplification of a Takamaka shared entity and of its connections with the objects persisted in the blockchain.}
\label{figure.shared_entity}
\end{figure}

@@

\begin{figure}[tp]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class SimpleSharedEntity
    <S extends PayableContract,O extends Offer<S>>
    extends PayableContract implements SharedEntity<S,O> {

@@

  @Override @View public final BigInteger sharesOnSaleOf(S shareholder) {
    return offers.stream()
      .filter(o -> o.seller == shareholder && o.isOngoing())
      .map(offer -> offer.sharesOnSale)
      .reduce(ZERO, BigInteger::add);
  }
}
    \end{lstlisting}
  \end{center}
  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
\end{figure}

>>>>

\section{A Generics Shared Entities Implementation}\label{sec:shared_entities}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figures/shared_entity}
\caption{Exemplification of a Takamaka shared entity and of its connections with the objects persisted in the blockchain.}
\label{figure.shared_entity}
\end{figure}

\begin{figure*}[ht]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public interface SharedEntity<S extends PayableContract, O extends Offer<S>>
  extends SharedEntityView<S> {

@@

    @View public boolean isOngoing() {
      return now() <= expiration;
    }
  }
}
    \end{lstlisting}
  \end{center}
  \caption{A simplified part of our shared entity interface.}\label{fig:shared_entity}
\end{figure*}

@@

@@

\begin{figure*}[tp]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class SimpleSharedEntity
    <S extends PayableContract,O extends Offer<S>>
    extends PayableContract implements SharedEntity<S,O> {

@@

  @Override @View public final BigInteger sharesOnSaleOf(S shareholder) {
    return offers.stream()
      .filter(o -> o.seller == shareholder && o.isOngoing())
      .map(offer -> offer.sharesOnSale)
      .reduce(ZERO, BigInteger::add);
  }
}
    \end{lstlisting}
  \end{center}
  \caption{A simplified part of our implementation of the shared entity interface.}\label{fig:simple_shared_entity}
\end{figure*}

### 2022-03-02 cluster_computing/040_solution.tex
title, abstract and introduction rewriting + biblio fix

\section{Fixing the Compilation of the Contract}\label{sec:fix}

@@

\begin{figure}[ht]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class TendermintValidators
   extends AbstractValidators<TendermintED25519Validator> {

@@

  @Override @FromContract(PayableContract.class) @Payable
  public void accept
    (BigInteger amount, TendermintED25519Validator buyer, 
     Offer<TendermintED25519Validator> offer) {
    super.accept(amount, buyer, offer);
  }
}
    \end{lstlisting}
  \end{center}
  \caption{The fixed code of the shared entity of the validators of a Hotmoka blockchain built over Tendermint.}\label{fig:solution}
\end{figure}

@@

  // synthetic bridge method
  public void accept(BigInteger,PayableContract,Offer)
   aload_0
   aload_1
   aload_2
   checkcast TendermintED25519Validator
   aload_3
   invokevirtual accept(BigInteger,TendermintED25519Validator,Offer)
   return
}
\end{lstlisting}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=.95\linewidth]{figures/solution}
\caption{Example of how the proposed solution works at run time when the \<accept> method is called by a \<Validator> or an \<Attacker>.}
\label{figure.solution}
\end{figure}

>>>>

\section{A Solution for Fixing the Compilation of the Contract}\label{sec:fix}

@@

\begin{figure*}[ht]
  \begin{center}
    \begin{lstlisting}[language=Takamaka]
public class TendermintValidators
   extends AbstractValidators<TendermintED25519Validator> {

@@

  @Override @FromContract(PayableContract.class) @Payable
  public void accept
    (BigInteger amount, TendermintED25519Validator buyer, 
     Offer<TendermintED25519Validator> offer) {
    super.accept(amount, buyer, offer);
  }
}
    \end{lstlisting}
  \end{center}
  \caption{The fixed code of the shared entity of the validators of a Hotmoka blockchain built over Tendermint.}\label{fig:solution}
\end{figure*}

@@

  // synthetic bridge method
  public void accept(BigInteger,PayableContract,Offer)
   aload_0
   aload_1
   aload_2
   checkcast TendermintED25519Validator
   aload_3
   invokevirtual accept
       (BigInteger,
        TendermintED25519Validator,
        Offer)
   return
}
\end{lstlisting}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.9\linewidth]{figures/solution}
\caption{Example of how the proposed solution works at run time when the \<accept> method is called by a \<Validator> or an \<Attacker>.}
\label{figure.solution}
\end{figure*}
